---
title: "Incorporating sketching into a typical single-cell analysis workflow"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
    %\VignetteIndexEntry{sketching_workflows}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    chunk_output_type: console
bibliography:
    bibliography.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  crop = NULL
)
library(BiocStyle)
```

## Introduction

In this vignette, we outline how geometric sketching can be incorporated into 
the steps of a typical single-cell analysis workflow. 

## Preparation

We start by loading the required packages and preparing an example data set. 

```{r load-pkg}
suppressPackageStartupMessages({
    library(sketchR)
    library(TENxPBMCData)
    library(scuttle)
    library(scran)
    library(scater)
    library(SingleR)
    library(cowplot)
    library(SummarizedExperiment)
    library(SingleCellExperiment)
    library(beachmat.hdf5)
    library(snifter)
    library(uwot)
    library(bluster)
    library(class)
})
```

We will use the PBMC3k data set from the `r Biocpkg("TENxPBMCData")` 
Bioconductor package for illustration. The chunk below prepares the data set
by calculating log-transformed normalized counts, finding highly variable 
genes and performing principal component analysis.

```{r process-data}
## Load data
pbmc3k <- TENxPBMCData(dataset = "pbmc3k")

## Set row and column names
colnames(pbmc3k) <- paste0("Cell", seq_len(ncol(pbmc3k)))
rownames(pbmc3k) <- uniquifyFeatureNames(
    ID = rowData(pbmc3k)$ENSEMBL_ID,
    names = rowData(pbmc3k)$Symbol_TENx
)

## Normalize and log-transform counts
pbmc3k <- logNormCounts(pbmc3k)

## Find highly variable genes
dec <- modelGeneVar(pbmc3k)
top_hvgs <- getTopHVGs(dec, n = 2000)

## Perform dimensionality reduction
set.seed(100)
pbmc3k <- runPCA(pbmc3k, subset_row = top_hvgs)
```

## Sketch cells

Next, we extract a subset of the cells using geometric sketching. We store 
information about which cells were sketched in the `colData` of `pbmc3k`. 

```{r sketch}
idx800gs <- geosketch(reducedDim(pbmc3k, "PCA"), 
                      N = 800, seed = 123)
pbmc3k$sketched <- seq_len(ncol(pbmc3k)) %in% idx800gs
table(pbmc3k$sketched)
```

We also generate a new set of highly variable genes, using only the sketched 
cells. This allows cells of different types to contribute more equally to the 
set of highly variable genes. 

```{r hvgs}
## Find highly variable genes
dec_sketched <- modelGeneVar(pbmc3k[, pbmc3k$sketched])
top_hvgs_sketched <- getTopHVGs(dec_sketched, n = 2000)
```

## Principal component analysis (PCA)

Here, we will illustrate how to apply PCA to the sketched cells, followed by 
projection of all cells onto the PCA representation obtained from the 
sketched cells. In this way, all cells are retained for further analysis, but 
only the sketched ones are used to determine the principal components. 

```{r pca}
#| fig.width: 8
#| fig.height: 5

set.seed(123)

# subset to sketched cells and run PCA, using the HVGs from the sketched subset
pbmc3k_sketched <- pbmc3k[, pbmc3k$sketched]
pca <- calculatePCA(pbmc3k_sketched, ncomponents = 30, 
                    subset_row = top_hvgs_sketched)

# store the PCA projection for the sketched cells as well as the weights for 
# the HVGs and the mean value for each HVG across the sketched cells
projection <- pca
rotation <- attr(pca, "rotation")
centers <- rowMeans(assay(pbmc3k_sketched[top_hvgs_sketched, ],
                          "logcounts"))

# center the full data set using the mean values from the sketched subset and 
# project all cells onto the extracted principal components
pca <- t(sweep(assay(pbmc3k[names(centers), ], "logcounts"), MARGIN = 1, 
               STATS = centers, FUN = "-")) %*% rotation

# add the PCA coordinates to pbmc3k
stopifnot(all(rownames(pca) == colnames(pbmc3k)))
reducedDim(pbmc3k, "PCA_sketch") <- as.matrix(pca)

# check that projections of sketched cells are retained
# ... via correlations
vapply(seq_len(ncol(projection)), function(i) {
    cor(projection[, i],
        reducedDim(pbmc3k, "PCA_sketch")[rownames(projection), i]
    )
}, NA_real_)

# ... via plotting
plot_grid(plotlist = lapply(seq_len(ncol(projection))[1:2], function(i) {
    ggplot(data.frame(
        Original = projection[, i],
        Reprojected = reducedDim(pbmc3k, "PCA_sketch")[pbmc3k$sketched, i]),
        aes(x = Original, y = Reprojected)) + 
        geom_point() +
        labs(title = paste0("PCA dimension ", i)) + 
        theme_bw()
}))
```

# t-SNE

Here, we illustrate how to project the full data set into a tSNE representation 
derived from the sketched cells. This workflow is also described in the 
`r Biocpkg("snifter")` vignette. Note that, unlike for PCA, for tSNE the 
projected coordinates of the sketched cells themselves may not be identical 
to their original coordinates in the tSNE representation, as the initial 
placement of a point in the embedding is obtained by averaging the coordinates
of its `k` nearest neighbors. Unless `k=1`, this initial placement will 
therefore be influenced by other sketched cells as well. If we want to make 
sure that the low-dimensional coordinates of the sketched cells are exactly 
those obtained by the original tSNE embedding, this can be achieved by only 
projecting the non-sketched cells, and concatenating these projections
with the original embeddings of the sketched cells. 

```{r tsne}
#| fig.width: 8
#| fig.height: 5

set.seed(123)

# extract tSNE embedding from sketched cells
tsne <- fitsne(
    reducedDim(pbmc3k[, pbmc3k$sketched], "PCA_sketch"), 
    n_jobs = 2L, random_state = 123L,
    perplexity = 30, n_components = 2L, 
    pca = FALSE)

# project all cells
proj <- project(
    x = tsne, 
    new = reducedDim(pbmc3k, "PCA_sketch"), 
    old = reducedDim(pbmc3k[, pbmc3k$sketched], "PCA_sketch"),
    perplexity = 30, k = 1L)
rownames(proj) <- colnames(pbmc3k)
reducedDim(pbmc3k, "TSNE_sketch") <- proj

# representations of sketched cells are approximately retained
# ... correlations
vapply(seq_len(ncol(tsne)), function(i) {
    cor(tsne[, i],
        reducedDim(pbmc3k, "TSNE_sketch")[pbmc3k$sketched, i]
    )
}, NA_real_)

# ... plotting
plot_grid(plotlist = lapply(seq_len(ncol(tsne)), function(i) {
    ggplot(data.frame(
        Original = tsne[, i],
        Reprojected = reducedDim(pbmc3k, "TSNE_sketch")[pbmc3k$sketched, i]),
        aes(x = Original, y = Reprojected)) + 
        geom_point() +
        labs(title = paste0("tSNE dimension ", i)) + 
        theme_bw()
}))
```

# UMAP

Similarly to tSNE, we can also project new data into an existing UMAP 
embedding, using the `r CRANpkg("uwot")` package. 

```{r umap}
#| fig.width: 8
#| fig.height: 5

set.seed(123)

# extract UMAP embedding from sketched cells
umap <- umap(
    X = reducedDim(pbmc3k[, pbmc3k$sketched], "PCA_sketch"), 
    n_components = 2, n_neighbors = 50,
    pca = NULL, scale = FALSE, 
    ret_model = TRUE, n_threads = 2L)

# project all cells
proj <- umap_transform(
    X = reducedDim(pbmc3k, "PCA_sketch"), 
    model = umap, 
    n_threads = 2L)
rownames(proj) <- colnames(pbmc3k)
reducedDim(pbmc3k, "UMAP_sketch") <- proj

# representations of sketched cells are approximately retained
# ... correlations
vapply(seq_len(ncol(umap$embedding)), function(i) {
    cor(umap$embedding[, i],
        reducedDim(pbmc3k, "UMAP_sketch")[pbmc3k$sketched, i]
    )
}, NA_real_)

# ... plotting
plot_grid(plotlist = lapply(seq_len(ncol(umap$embedding)), function(i) {
    ggplot(data.frame(
        Original = umap$embedding[, i],
        Reprojected = reducedDim(pbmc3k, "UMAP_sketch")[pbmc3k$sketched, i]),
        aes(x = Original, y = Reprojected)) + 
        geom_point() +
        labs(title = paste0("UMAP dimension ", i)) + 
        theme_bw()
}))
```

# Clustering

Also clustering can be beneficial to apply only to the sketched cells, followed
by prediction of cell type labels using, e.g., k-nearest neighbors. Following 
a similar reasoning as for tSNE above, unless we use `k=1` in the k-nearest 
neighbor prediction, the predicted cluster labels of the sketched cells may 
not be the same as the ones obtained directly from the clustering. Here we 
cluster the sketched cells using functions from the `r Biocpkg("bluster")`
package. 

```{r clustering}
# subset to only the sketched cells
# note that since the PCA_sketch embedding was obtained by applying PCA to the 
# sketched cells only, subsetting like this is equivalent to applying PCA 
# again to the sketched cells only (see above for an illustration)
pca_sketch <- reducedDim(pbmc3k[, pbmc3k$sketched], "PCA_sketch")

set.seed(123)

# run clustering on the sketched cells
clst <- as.character(clusterRows(
    pca_sketch,  
    BLUSPARAM = SNNGraphParam(
        k = 10, type = "rank", 
        num.threads = 2L, 
        cluster.fun = "leiden", 
        cluster.args = list(resolution = 0.15))))

# predict labels of all cells using 1-nearest neighbors and add cluster labels
# to pbmc3k
preds <- knn(train = pca_sketch, 
             test = reducedDim(pbmc3k, "PCA_sketch"), 
             cl = clst, k = 1) |> as.character()
pbmc3k$clusters <- preds

# check that labels are retained for sketched cells
all(clst == pbmc3k$clusters[pbmc3k$sketched])
```

# Session info

```{r session-info}
sessionInfo()
```

