---
title: "Subsampling single-cell data sets with sketchR"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
    %\VignetteIndexEntry{sketchR}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    chunk_output_type: console
bibliography:
    bibliography.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  crop = NULL
)
library(BiocStyle)
```

## Introduction

This vignette showcases the main functionalities of the `sketchR` package, 
and illustrates how it can be used to generate a subsample of a dataset using 
the geometric sketching/subsampling algorithms and implementations proposed by 
@Hie2019-geosketch and @Song2022-scsampler, as well as create a set of 
diagnostic plots. 

## Preparation

We start by loading the required packages and preparing an example data set. 

```{r setup}
suppressPackageStartupMessages({
    library(sketchR)
    library(TENxPBMCData)
    library(scuttle)
    library(scran)
    library(scater)
    library(SingleR)
    library(celldex)
    library(cowplot)
    library(SummarizedExperiment)
    library(SingleCellExperiment)
})
```

We will use the PBMC3k data set from the `r Biocpkg("TENxPBMCData")` 
Bioconductor package for illustration. The chunk below prepares the data set
by calculating log-transformed normalized counts, finding highly variable 
genes, performing dimensionality reduction and predicting cell type labels 
using the `r Biocpkg("SingleR")` package.

```{r}
## Load data
pbmc3k <- TENxPBMCData::TENxPBMCData(dataset = "pbmc3k")

## Set row and column names
colnames(pbmc3k) <- paste0("Cell", seq_len(ncol(pbmc3k)))
rownames(pbmc3k) <- scuttle::uniquifyFeatureNames(
    ID = SummarizedExperiment::rowData(pbmc3k)$ENSEMBL_ID,
    names = SummarizedExperiment::rowData(pbmc3k)$Symbol_TENx
)

## Normalize and log-transform counts
pbmc3k <- scuttle::logNormCounts(pbmc3k)

## Find highly variable genes
dec <- scran::modelGeneVar(pbmc3k)
top.hvgs <- scran::getTopHVGs(dec, n = 2000)

## Perform dimensionality reduction
set.seed(100)
pbmc3k <- scater::runPCA(pbmc3k, subset_row = top.hvgs)
pbmc3k <- scater::runTSNE(pbmc3k, dimred = "PCA")

## Predict cell type labels
ref_monaco <- celldex::MonacoImmuneData()
assay(pbmc3k, "logcounts") <- as.matrix(assay(pbmc3k, "logcounts"))
pred_monaco_main <- SingleR::SingleR(test = pbmc3k, ref = ref_monaco, 
                                     labels = ref_monaco$label.main)
pbmc3k$labels_main <- pred_monaco_main$labels

dim(pbmc3k)
```

## Subsampling

The `geosketch()` function performs geometric sketching by calling the 
`geosketch` [python package](https://github.com/brianhie/geosketch). 
The output is a vector of indices that can be used to subset the full 
dataset. The provided seed will be propagated to the python code to 
achieve reproducibility. 

```{r}
idx800gs <- geosketch(SingleCellExperiment::reducedDim(pbmc3k, "PCA"), 
                      N = 800, seed = 123)
head(idx800gs)
length(idx800gs)
```

Similarly, the `scsampler()` function calls the `scSampler` 
[python package](https://github.com/SONGDONGYUAN1994/scsampler) to 
perform subsampling. 

```{r}
idx800scs <- scsampler(SingleCellExperiment::reducedDim(pbmc3k, "PCA"), 
                       N = 800, seed = 123)
head(idx800scs)
length(idx800scs)
```

To illustrate the result of the subsampling, we plot the tSNE 
representation of the original data as well as the two subsets (using the 
tSNE coordinates derived from the full dataset).

```{r, fig.width = 10, fig.height = 8}
cowplot::plot_grid(
    scater::plotTSNE(pbmc3k, colour_by = "labels_main"),
    scater::plotTSNE(pbmc3k[, idx800gs], colour_by = "labels_main"),
    scater::plotTSNE(pbmc3k[, idx800scs], colour_by = "labels_main")
)
```

We can also illustrate the relative abundance of each cell type in the 
full data and in the subsets, respectively. 

```{r, fig.width = 6, fig.height = 8}
compareCompositionPlot(SummarizedExperiment::colData(pbmc3k), 
                       idx = list(geosketch = idx800gs,
                                  scSampler = idx800scs), 
                       column = "labels_main")
```

## Diagnostic plots

`sketchR` provides a convenient function to plot the Hausdorff distance 
between the full dataset and the subsample, for a range of sketch 
sizes. 

```{r}
hausdorffDistPlot(mat = SingleCellExperiment::reducedDim(pbmc3k, "PCA"), 
                  Nvec = c(400, 800, 2000), seed = 123, 
                  Nrep = 3, methods = c("geosketch", "scsampler", "uniform"))
```

## Session info

```{r}
sessionInfo()
```

## References
